Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2024-07-31 EvolutionaryRescue1.tex
revise latest changes by RS

\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug-resistant and the other drug-sensitive, can evade extinction by cells switching between the two phenotypes through epigenetic mutations. They found that even when drug-resistant cells are barely viable, the epimutations guarantee evolutionary rescue. Evolutionary rescue in a single step, where an initially declining population has to acquire a mutation to survive extinction after a sudden environmental change, has been studied by \citet{orr2008population,orr2014population}. They analyzed a model where the mutant strain is present in small numbers at the onset of therapy. They concluded that this can significantly enhance the population's probability of survival. \citet{iwasa2003evolutionary,iwasa2004evolutionary} analysed the probability that a population under strong selective pressure can survive extinction   through succesive mutations with the help of multitype branching processes.

Most models focus on the probability that at most one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.  Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability}, who tested their model's predictions with data from yeast and bacteria experiments.

@@

We use multi-type branching-process theory to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_s$ in three distinct scenarios (Appendix~\ref{sec:appendix-surv-prob}). 
Substituting these into $N^*=1/p_s$, we find approximations for the threshold tumor size, $N^*$. 
In these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then it is likely that evolutionary rescue will occur through a direct resistance mutation in a sensitive cell without aneuploidy playing a role in the adaptive dynamics, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_s}}{v\lambda_s}  \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_s}/\left(v\lambda_s\right)$ is the ratio of the rate at which sensitive cells decrease in number and the rate at which they are mutating. Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can be approximated by one of the following scenarios, depending on $\Delta_a T^*$, which represents the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_s}}{u\lambda_s} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to the results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the scenario in which the treatment still effectively kills aneuploid cells but not as quickly as the sensitive cells. 
In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly 1.
In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease in the tumor population size, allowing more opportunities to produce resistant mutants. 
In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.
Notably, in this scenario the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all scenarios, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second scenario where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A). If the mutation rate $v$ increases the tumor threshold size for tolerant aneuploids will decrease $\sim v^{-1}$ while the threshold size for stationary aneuploids will decrease $\sim v^{-\frac{1}{2}}$. The growth rate for which tolerant aneuploids can offer a viable path to rescuing the cancer cell population is in the interval $\left(-u\lambda_a,-1/T^*\right)$ which will become smaller as $v$ increases. As a result, for any value of $\Delta_a\ll-\left(T^*\right)^{-1}$ increasing the mutation rate will improve probability of evolutionary rescue of the trajectory $sensitive\rightarrow aneuploid\rightarrow mutant$ until the threshold $\Delta_a=-\left(T^*\right)^{-1}$ is reached.
% TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4 ok

@@

Our results quantitatively suggest that aneuploidy may play an important role in tumor adaptation to anti-cancer drugs when the tumor size is small or intermediate. Large tumors are predicted to adapt to anti-cancer drugs through direct mutation. In contrast, smaller tumors are predicted to become resistant either directly by aneuploidy or by a resistance mutation occurring in aneuploid cells that serve as an evolutionary ``stepping stones'' (\Cref{rescue_prob}). Thus, therapies that increase the rate of aneuploidy in tumors to combat cancer may also promote drug resistance.

>>>>

\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug-resistant and the other drug-sensitive, can evade extinction by cells switching between the two phenotypes through epigenetic mutations. They found that even when drug-resistant cells are barely viable, the epimutations guarantee evolutionary rescue. Evolutionary rescue in a single step, where an initially declining population has to acquire a mutation to survive extinction after a sudden environmental change, has been studied by \citet{orr2008population,orr2014population}. They analyzed a model where the mutant strain is present in small numbers at the onset of therapy. They concluded that this can significantly enhance the population's probability of survival.

Most models focus on the probability that at most one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue. Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability}, who tested their model's predictions with data from yeast and bacteria experiments. \citet{iwasa2003evolutionary,iwasa2004evolutionary} used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction via succesive mutations.

@@

We use multi-type branching-process theory to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_s$ in three distinct scenarios (Appendix~\ref{sec:appendix-surv-prob}). 
Substituting these into $N^*=1/p_s$, we find approximations for the threshold tumor size, $N^*$. 
In these approximations, an important quantity is $T^* = \sqrt{\lambda_m / 4 v \lambda_a^2 \Delta_m}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then it is likely that evolutionary rescue will occur through a direct resistance mutation in a sensitive cell without aneuploidy playing a role in the adaptive dynamics, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_s}}{v\lambda_s}  \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_s}/\left(v\lambda_s\right)$ is the ratio of the rate at which sensitive cells decrease in number and the rate at which they are mutating. Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can be approximated by one of the following scenarios, depending on $\Delta_a T^*$, which represents the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_s}}{u\lambda_s} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to the results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the scenario in which the treatment still effectively kills aneuploid cells but not as quickly as the sensitive cells. 
In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly 1.
In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease in the tumor population size, allowing more opportunities to produce resistant mutants. 
In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.
Notably, in this scenario the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all scenarios, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second scenario where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A). 
The tumor threshold size decreases with the mutation rate in the first and second scenarios: $N_a^*$ is proportional to $1/v$ in the first scenario (tolerant aneuploids) and to $\sqrt{1/v}$ in the second scenario (stationary aneuploids). Furthermore, the growth rate $\Delta_a<0$ that allows tolerant aneuploids to rescue the tumor is between $-u\lambda_a$ and  $-1/T^*$, which is proportional to $-\sqrt{v}$. Thus, increasing the mutation rate $v$ will decrease the tumor threshold size $N^*_a$, making evolutionary rescue more likely, but only until $T^*$ decerases to a point where  $\Delta_a =1/T^*$.

@@

Our results quantitatively suggest that aneuploidy may play an important role in tumor adaptation to anti-cancer drugs when the tumor size is small or intermediate. Large tumors are predicted to adapt to anti-cancer drugs through direct mutation. In contrast, smaller tumors are predicted to become resistant either directly by aneuploidy or by a resistance mutation occurring in aneuploid cells that serve as evolutionary ``stepping stones'' (\Cref{rescue_prob}). Thus, therapies that increase the rate of aneuploidy in tumors to combat cancer may also promote drug resistance.
